---
title: "Workshop 4"
author: "Aiden Bennett"
date: "2026-05-28"
---

--- Workshop 04 - Developers Toolbox and AI ---

Today's workshop focuses on AI integration into R and setting up GitHub Copilot and chattr.

-- Installing and loading chattr --
```{r}
install.packages("ellmer")
install.packages("chattr")
```

An API key is also required to run these programs, I chose to use Google Gemini via Google AI Studio. Once I had the key, I needed to store the key in R using the usethis package. It is stored in the .Renviron file and was connected to chattr through the .Rprofile.

After each file was loaded R needed to be restarted.

I mapped the chattr app to the Alt+Shift+G key bind to make opening the program easier and asked it some simple questions to ensure it was working as intended. 

I asked it how to load the built-in data set cars and create a ggplot using some of its variables. This was the outputted code.
```{r}
library(ggplot2)

data(cars)

ggplot(cars, aes(x = speed, y = dist)) +
  geom_point()
```

-- Conditional Logic in R --

An example of an if/else loop in basic R
```{r}
sst <- 30.2
if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}
```

This code would output one message based on the value of SST after checking its value against both potential outputs.

An alternative using tidyverse exists that can combine the functions into one.
```{r}
library(tidyverse)

#Example of coral data
coral_monitoring <- tibble(
    site = c("Site_A", "Site_B", "Site_C"),
    depth_m = c(12, 35, 8)
)

# Creating the if_else
coral_monitoring <- coral_monitoring %>%
  mutate(zone = if_else(depth_m > 30, "Deep Reef", "Shallow Reef"))
```

This code instead creates a new output, coral_monitoring, that then gets updated using the mutate function to add a new column zone that then has its values updated using the if_else function.

If you have more than 2 options then a case_when function can also be used.
```{r}
coral_monitoring <- coral_monitoring %>% 
  mutate(reef_category = case_when(
    depth_m < 10  ~ "Lagoon / Flats",
    depth_m <= 30 ~ "Crest / Slope",
    depth_m > 30  ~ "Mesophotic / Deep",
    TRUE          ~ "Unclassified" # Catch-all remainder
  ))
```

I used the chattr app to make a new section of code with set salinity values and a mutate/case_when function to combine them into a new data chart. Below is the output.
```{r}
library(dplyr)
library(tibble)

marine_stations <- tibble(
  salinity = c(35, 28, 32, 12)
) %>%
  mutate(
    environment_type = case_when(
      salinity < 15 ~ "Estuarine",
      salinity >= 15 & salinity <= 30 ~ "Brackish",
      salinity > 30 ~ "Marine"
    )
  )
```

The AI made the whole code in one section, which further simplified the provided examples.

-- Automation --

Base R can create for-loops
```{r}
for (year in 2020:2024) {
  print(paste("Processing climate data for year:", year))
}
```

R can also make loops using index sequences
```{r}
transect_lengths <- c(50, 100, 25, 75)

for (i in seq_along(transect_lengths)) {
  print(paste("Transect number", i, "measures", transect_lengths[i], "metres."))
}
```

Tidyverse once again can be used to make useful functions, in this case map functions using the purrr package

map () - returns a flexible list structure
map_dbl() - returns a vector of decimals/numeric values
map_chr() - returns a vector of text strings/characters
map_df() - returns a combined data frame/tibble

A standard for loop and the purrr equivalent
```{r}
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas)) # Must build an empty container first

for(i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])
}

library(purrr)
# Map directly and define your expected output type explicitly
mapped_results <- map_dbl(site_areas, sqrt)
```

Both sections will output the same results

With chattr, I used this prompt. "I have a list of three vectors containing fish count data. Write a for-loop to calculate the mean of each, and then show me the exact parallel way to do it using purrr's map_dbl function.
```{r}
library(purrr)

fish_counts <- list(
  river_a = c(10, 12, 8, 15, 11),
  river_b = c(20, 18, 22, 19),
  river_c = c(5, 7, 6)
)

# For-loop
mean_counts_for_loop <- numeric(length(fish_counts))
names(mean_counts_for_loop) <- names(fish_counts)

for (i in seq_along(fish_counts)) {
  mean_counts_for_loop[i] <- mean(fish_counts[[i]])
}

mean_counts_for_loop

# purrr::map_dbl
mean_counts_purrr <- map_dbl(fish_counts, mean)

mean_counts_purrr
```

-- Creating Custom Functions --
When creating custom functions, it requires a Name, Arguments/Inputs, and a Body/Logic

Here is a provided example of a function
```{r}
# Function Definition
calculate_coral_mortality <- function(initial_count, surviving_count) {
  
  # Logic safety switch using our conditional tools!
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  
  mortality_rate <- (initial_count - surviving_count) / initial_count
  return(mortality_rate)
}

# Utilizing your custom function
calculate_coral_mortality(initial_count = 120, surviving_count = 84) #Result is 0.3
```

I needed to create a custom function called convert_temp_c_to_f that accepts one argument as temp in Celsius. It needs to perform the conversion and a safety check using an if argument to check if value is below absolute zero (-273.15 degrees C).
```{r}
convert_temp_c_to_f <- 
  function(temp_c) {
    if (temp_c < -273.15) {
      message("Input temperature is below absolute zero")
      return(NA_real_)
    } else {
      temp_f <- (temp_c * 9/5) + 32
      return(temp_f)
    }
  }

#Tests to see if function works
convert_temp_c_to_f(40) # Returns 104
convert_temp_c_to_f(-300) # Returns NA
```

AI integration and logic can help to simplify R functions and allow for more accessibility to complex coding.