---
title: "Workshop 1"
author: "Aiden Bennett"
date: "2026-05-20"
---

### --- Introduction ---
This document contains my workings through the first module of MB5370, with annotations explaining the functions and what some outputs represent.

### --- Directory ---
C:/Documents/Masters Docs General/MB5370 - Techniques in Marine Science 1/Module_1.Rmd

### --- Workshop 01 - Introduction ---
This workshop is the first introduction to coding and goes over the basic functions of R and how different types of data can be stored and utilized.

### -- Getting Started --
```{r message=FALSE, warning=FALSE}
2 + 1 # Addition
1:30 # A list from 1 to 30
6 * 2 # Multiplication
```

### -- Variables and Assignment --
```{r message=FALSE, warning=FALSE}
age <- 25 # Creating a variable and setting its value manually
first_name <- 'Bill'

# Can update a variable using functions from above
age + 1
age + age

# Multiple values can be combined at once and made into new variables at the same time
sum <- sum(15, 25.1, 20.25) #60.35
num <- 30.41
sum + num # 90.76
```

### -- Functions --
```{r message=FALSE, warning=FALSE}

# Rounding up and down to the whole number
years_old <- 25.7
round(years_old) # 26
floor(years_old) # 25

# Rounding with specified decimal points
years_old <- 25.765
round(years_old, 2) # 25.76

# The paste function allows for multiple variables to be combined into one output
name <- 'Aiden'
my_age <- '26 years old'
paste(name, my_age, sep = " is ", collapse = NULL, recycle0 = FALSE) # Will combine the two and put is in between
```

### -- Misconceptions --
```{r message=FALSE, warning=FALSE}

# When adjusting values it is important to know the load order of the functions
grade <- 55
total <- grade + 10
print(total) # will be 65
grade <- 90
print(total) # will still be 65 as total has not been updated
total <- grade + 10
print(total) # now is 100

p <- 2
z <- 5
out <- p * z # Should be 10
print(out) # is 10

x <- 1
is.character(x) # FALSE
is.numeric(x) # TRUE

my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro", # Missing comma at end
             "cinco")
print(my_quiz) # changed quis to quiz
str(my_quiz)
length(my_quiz) #should be length, not len
```

### -- Data Types --
```{r message=FALSE, warning=FALSE}
# Every variable has its own data type, important for later funcitons that can only use certain types of data
number <- 17
class(number) # Numeric
class(name) # Character
int <- 3L # needs to be not just a number or letter
class(int) # Integer
log <- 12 > 11
class(log) # Logical

#Data can also be changed using as.numeric/as.character etc.
```

### -- Data Structures --
```{r message=FALSE, warning=FALSE}
#Matricies and larger data sets with have complex structures that need to be navigated
y <- c(1, 2, 3) # makes a list showing 1, 2, 3
z <- c("Sarah", "Tracy", "Jon") # makes a list of names
class(y) # Numeric
class(z) # Character
```

### -- Data Frames --
```{r message=FALSE, warning=FALSE}
# A data frame is the way that most data sets will be displayed and show values of multiple variables at the same time
my_dataframe <- data.frame(no = c(1,2,3), c("Tracey", "John", "Pete"), c(TRUE, FALSE, TRUE))
my_dataframe # makes a 3x3 data frame 
str(my_dataframe)

my_dataframe$no = as.factor(my_dataframe$no)
str(my_dataframe)
```

### -- Packages --
```{r message=FALSE, warning=FALSE}
# Need to be loaded before any functions can be used, and installed if not already on device using install.packages()
library(tidyverse)
library(ggplot2)
```

### --- Exercise ---
How to load a .csv file and plot

1. Read the .csv data set into R from the directory and the read.csv function.
2. Once the table is loaded, print(file_name) will show what is in the table and the names of all of the variables.
3. With the names of the variables, ggplot(file_name, aes(species_name, variable 2)) or a simpler plot(file_name$species_name) can be done.
