---
output:
  word_document: default
---
MB5370 - Introduction of Programming
Aiden Bennett
May 2026

Directory:
C:/Documents/Masters Docs General/MB5370 - Techniques in Marine Science 1/Module_1.Rmd

--- Introduction ---
This document contains my workings through the first module of MB5370, with annotations explaining the functions and what some outputs represent.

--- Workshop 01 - Introduction ---
This workshop is the first introduction to coding and goes over the basic functions of R and how different types of data can be stored and utilized.

-- Getting Started --
```{r message=FALSE, warning=FALSE}
2 + 1 # Addition
1:30 # A list from 1 to 30
6 * 2 # Multiplication
```

-- Variables and Assignment --
```{r message=FALSE, warning=FALSE}
age <- 25 # Creating a variable and setting its value manually
first_name <- 'Bill'

# Can update a variable using functions from above
age + 1
age + age

# Multiple values can be combined at once and made into new variables at the same time
sum <- sum(15, 25.1, 20.25) #60.35
num <- 30.41
sum + num # 90.76
```

-- Functions --
```{r message=FALSE, warning=FALSE}

# Rounding up and down to the whole number
years_old <- 25.7
round(years_old) # 26
floor(years_old) # 25

# Rounding with specified decimal points
years_old <- 25.765
round(years_old, 2) # 25.76

# The paste function allows for multiple variables to be combined into one output
name <- 'Aiden'
my_age <- '26 years old'
paste(name, my_age, sep = " is ", collapse = NULL, recycle0 = FALSE) # Will combine the two and put is in between
```

-- Misconceptions --
```{r message=FALSE, warning=FALSE}

# When adjusting values it is important to know the load order of the functions
grade <- 55
total <- grade + 10
print(total) # will be 65
grade <- 90
print(total) # will still be 65 as total has not been updated
total <- grade + 10
print(total) # now is 100

p <- 2
z <- 5
out <- p * z # Should be 10
print(out) # is 10

x <- 1
is.character(x) # FALSE
is.numeric(x) # TRUE

my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro", # Missing comma at end
             "cinco")
print(my_quiz) # changed quis to quiz
str(my_quiz)
length(my_quiz) #should be length, not len
```

-- Data Types --
```{r message=FALSE, warning=FALSE}
# Every variable has its own data type, important for later funcitons that can only use certain types of data
number <- 17
class(number) # Numeric
class(name) # Character
int <- 3L # needs to be not just a number or letter
class(int) # Integer
log <- 12 > 11
class(log) # Logical

#Data can also be changed using as.numeric/as.character etc.
```

-- Data Structures --
```{r message=FALSE, warning=FALSE}
#Matricies and larger data sets with have complex structures that need to be navigated
y <- c(1, 2, 3) # makes a list showing 1, 2, 3
z <- c("Sarah", "Tracy", "Jon") # makes a list of names
class(y) # Numeric
class(z) # Character
```

-- Data Frames --
```{r message=FALSE, warning=FALSE}
# A data frame is the way that most data sets will be displayed and show values of multiple variables at the same time
my_dataframe <- data.frame(no = c(1,2,3), c("Tracey", "John", "Pete"), c(TRUE, FALSE, TRUE))
my_dataframe # makes a 3x3 data frame 
str(my_dataframe)

my_dataframe$no = as.factor(my_dataframe$no)
str(my_dataframe)
```

-- Packages --
```{r message=FALSE, warning=FALSE}
# Need to be loaded before any functions can be used, and installed if not already on device using install.packages()
library(tidyverse)
library(ggplot2)
```

--- Exercise ---
How to load a .csv file and plot

1. Read the .csv data set into R from the directory and the read.csv function.
2. Once the table is loaded, print(file_name) will show what is in the table and the names of all of the variables.
3. With the names of the variables, ggplot(file_name, aes(species_name, variable 2)) or a simpler plot(file_name$species_name) can be done.


--- Workshop 02 - Visualization ---
The second workshop introduces plotting using the ggplot2 package from tidyverse, going over types of plots and the different ways that they can be customized using embedded functions.

-- Load packages --
```{r message=FALSE, warning=FALSE}
# Loads ggplot2 at same time
library(tidyverse)
```

-- Plotting practice with ggplot2 --
```{r message=FALSE, warning=FALSE}
# ggplot2 is a great package for making more complex charts and removes redundancy that the base version of R plotting usually has
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) # A simple dot chart

# Changing point colour
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) # Sets the colours based on the variable class

# Changing point size
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class)) # Changes size as above with the class variable

# Changing point transparency
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) # Gives each class its own transparency level

# Changing point shape
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class)) # Assign's each class its own shape

# Manually set colour
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") # Manually makes all dots blue
```

-- Facet and panel plots --
```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
# This function gives each subplot its own panel to separate the values of class from one another

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
#This function instead combines the variables of drv and cyl together using an ~ to show relation between the variables

?facet_wrap # nrow defines the number of rows, ncol defines the number of columns. Scales and space define limits on the layout and ensure that each panel is equal and proportional
```

-- Fitting simple lines --
```{r message=FALSE, warning=FALSE}
# Using geom_point
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
# geom_point only maps dot points, and often requires additional adjustments of aesthetics or graph styles to be readable

# Using geom_smooth
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
# geom_smooth only makes a line, and like geom_point will need more adjustments. It does however have an area around it that shows where points are more or less concentrated 

# Changing the line style
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
# This plot has three different types of lines based on values of drv, which can help distinguish what trend each line is representing

#Set group aesthetic
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
# This plot resets the lines from above and instead just draws the three lines

# Change group colour
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )
# Using color it sets each line to a unique color based on the values of a specified variable, in this case it is still drv

# Points and lines together
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
# Going back to one line, we combine the line and dot charts into one plot together and visualize the relationship between them

# Same but simpler
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
# This produces the same plot as the previous but does not repeat the aes code in the geom functions

# Style points
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
# Only changes the style of the points, in this case changing their colour

# Subset data during plotting
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
# Filter sets the data into specified categories then only plots the specified subset
```

--- Exercise ---
1. For a line chart use geom_line, geom_boxplot for a boxplot, geom_histogram for a histogram, and geom_area for an area chart. These will be the simplest options and can be easily modified.

2. These plots will show a dot chart with a best-fit line going through, evidenced by geom_point and geom_smooth. They will look the same as the mapping sections are the same in both plots, only placed in different locations within the code.
```{r message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

-- Plotting Statistics --
```{r message=FALSE, warning=FALSE}
# Plots from diamonds data set in ggplot
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut)) # Makes a bar chart

# Changing defaults in ggplot2
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
# This adjusts the default statistic (count) to the raw value of the variable

# Alternative to overridding defaults
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
# This plot specifically moves the values to a proportional graph

# Stat summary
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
# stat_summary can visualize transformations of the data
```

-- Aesthetic adjustments --
```{r message=FALSE, warning=FALSE}
# Color and fill
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) 
# Colors the outline

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
# Colors the inside of the bar

# Clarity
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
# Makes subsections in the bar that are colored by the variable clarity, a form of a subset

# Alter transparency
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
# Transparency is another way to show subsets in the bars

# Color outlines w/o fill
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
# Removes the fill of all of the bars and colors their outlines

# Fill
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
# Using this function it makes all of the bars the same height while still coloring them

# Dodge
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
# Dodge positions the overlapping values next to each other instead of on top

# Jitter
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
# Jitter is only usable in dot plots and adds randomization to the points to stop them from being combined in the plots
```

-- Layered grammar of plotting --
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <FACET_FUNCTION>
  
The location of each important factor in a ggplot, and can be used as a base template for most plotting. When multiple variables are being used then it will need to be adjusted but this is good enough for most basic requirements of ggplot.